---
title: "BCM Datathon Team 17 Data Cleaning"
author: "Andrew Zimolzak, MD, MMSc"
date: 'April, 2022'
output:
  pdf_document: default
  html_document: default
---

# Contents

1. Conclusions up front
2. Import patient visits
3. Import demographics
4. Import clinical notes
5. Make analytic data set
6. Prelim results

# Conclusions so far

There are more women (71 vs 63%) and Latino patients (36 vs 22%) represented among notes containing any negative descriptor, compared to notes without a negative descriptor. There are slightly fewer black patients (46 vs 50%) represented among notes with any negative descriptor.

- More women when negative descriptors = 1, 2 but fewer when = 3, 4.
- More Latino patients when negative descriptors = 1 or 2, but fewer when = 3, 4.
- Fewer black patients when negative descriptors = 1, more when = 2, 3, 4, 5, 6.

No adjustment yet for covariates. To see quantitative association results, go to "RESULTS" section (probably the final 2 pages of the PDF).


```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(dplyr)
library(lubridate)
library(data.table)
library(knitr)
library(tableone)
# library(readxl)

source(file=here("functions.R"))
```

```{r import, echo=FALSE}

demog = filename2df('Patient Demographics 17 20220326.txt')  # 0.6 MB
# comor = filename2df('Patient Comorbidities 17 20220327.txt')  # 2.7 MB
edadm = filename2df('Patient ED Admissions 17 20220326.txt')  # 2.9 MB
# edpro = filename2df('Patient ED DX Problem List 17 20220327.txt')  # 19 MB
# medad = filename2df('Patient Meds Admin Times 17 20220327.txt')  # 93 MB
# order = filename2df('Patient Orders 17 20220327.txt')  # 86 MB
notes = read.csv(here('onedrive', 'Working', 'DATA', 'Progress Notes File.csv'),
                 sep=",", stringsAsFactors = FALSE)
```

```{r tidy encounters, echo=FALSE}
edadm %>%
  mutate(
    ed_adm = ymd_hms(EMERGENCY_ADMISSION_DTTM),
	  ed_dis = ymd_hms(ED_DEPARTURE_DTTM),
	  inp_adm = ymd_hms(HOSP_ADMSN_TIME),
	  inp_dis = ymd_hms(HOSP_DISCH_TIME),
  ) %>%
  rename(cc10 = First.Chief.Compliant.ICD10, cc9 = First.Chief.Compliant.ICD9) %>%
  select(
	  PAT_ID, ED.Disposition, Patient.Class, cc10, cc9,
	  ed_adm, ed_dis, inp_adm, inp_dis
  ) %>%
  mutate(
	  adm_adm = (inp_adm - ed_adm) / ddays(),
	  dis_dis = (inp_dis - ed_dis) / ddays(),
	  ed_dur = (ed_dis - ed_adm) / ddays(),
	  inp_dur = (inp_dis - inp_adm) / ddays(),
	  er_vs_inp = inp_dur - ed_dur
  ) -> ed_encounter_tidy
```

# Patient visits (descriptive stats)

## Encounter class

```{r}
table(ed_encounter_tidy$Patient.Class) %>% kable()
```

## ED disposition

```{r}
table(ed_encounter_tidy$ED.Disposition) %>% kable()
```

## Skip this section (Andy's notes)

```{r crosstab, include=FALSE}
table(ed_encounter_tidy$ED.Disposition, ed_encounter_tidy$Patient.Class) %>% kable()

```

My observations:

-   disposition=Discharge and class=Emergency: common but not perfectly?
-   d=Admit c=Inpatient: common
-   d=Observation c=Observation: somewhat common
-   d=Observation c=Inpatient: even more common (surprisingly)
-   d=Admit class=Observation: common
-   c=Outpatient: uncommon in general
-   disposition=Admit class=Emergency: rare

The main dispositions that seem to matter:

-   Discharge
-   Admit
-   Observation
-   Transfer

We could make a more limited table of 4 dispo X 3 class. This is all possibly because we requested ER discharge to home and not admit?

```{r inspect me, include=FALSE}
ed_encounter_tidy %>%
  select(-PAT_ID, -cc10, -cc9) -> sample_me

sample_me %>%
  slice_sample(n=20) %>%
  arrange(Patient.Class, ED.Disposition) %>%
  transmute(
    class = tolower(substr(Patient.Class, 1, 3)),
    dispo = tolower(substr(ED.Disposition, 1, 3)),
    # er1 = ed_adm,
    # er2 = ed_dis,
    # in1 = inp_adm,
    # in2 = inp_dis,
    er1_to_inp1 = round(adm_adm, 2),
    er2_to_inp2 = round(dis_dis, 2),
    los_er = round(ed_dur, 2),
    los_inp = round(inp_dur, 2),
    diff_los = round(er_vs_inp, 2)
  ) %>%
  kable()
```


Looks like if dispo = discharge, then ER departure time = Hospital discharge time.

## Distribution of encouters per patient ID

```{r, message=FALSE}
ed_encounter_tidy %>%  count(PAT_ID) -> pat_id_counts
qplot(x=n, data=pat_id_counts) +  scale_x_log10()
```

## Skip this section (Andy's notes)

Why do some have only 1 encounter?

```{r echo=FALSE}
left_join(ed_encounter_tidy, pat_id_counts, by='PAT_ID') %>%
  filter(Patient.Class != 'Outpatient') %>%
  filter(
    ED.Disposition == 'Admit' |
      ED.Disposition == 'Discharge' |
      ED.Disposition == 'Observation' |
      ED.Disposition == 'Transfer to Another Facility'
  ) %>%
  mutate(
    multiple = case_when(
      n > 1 ~ 'mult',
      TRUE ~ 'single'
    )
  ) -> ed_enc_w_counts
```

```{r include=FALSE}
table(ed_enc_w_counts$multiple, ed_enc_w_counts$Patient.Class) %>% kable()
```

```{r include=FALSE}
table(ed_enc_w_counts$multiple, ed_enc_w_counts$ED.Disposition) %>% kable()
```

```{r include=FALSE}
ed_enc_w_counts %>% filter(multiple == 'mult') -> m
ed_enc_w_counts %>% filter(multiple == 'single') -> s
```

```{r crosstab single, include=FALSE}
table(s$Patient.Class, s$ED.Disposition) %>% kable()
```

```{r crosstab mult, include=FALSE}
table(m$Patient.Class, m$ED.Disposition) %>% kable()
```

```{r inspect mult, include=FALSE}
m %>% 
  select(PAT_ID, Patient.Class, ED.Disposition,
         ed_adm, ed_dis, inp_adm, inp_dis, n) %>%
  filter(n < 5) %>%
  arrange(PAT_ID, ed_adm) %>%
  head(25) %>%
  kable() -> inspect_multiples

# inspect_multiples
```

Inspect singles by themselves.

```{r inspect single, include=FALSE}
s %>% 
  select(PAT_ID, Patient.Class, ED.Disposition,
         ed_adm, ed_dis, inp_adm, inp_dis, er_vs_inp, n) %>%
  arrange(PAT_ID, ed_adm) %>%
  head(25) %>%
  kable() -> inspect_singles

# inspect_singles
```

2022-04-13 meeting notes.

ED admiss and ED discharge is literally when they came/went from ED. Whereas, inpatient admit/dis date and time are about when they *first showed up to hospital* no matter what part of hospital. Many of the obs people will expect to have inp discharge time which is *after* the ED disch time. But not *all.* Occasional data error or whatever.

Also some things come thru diff in Slicer vs Caboodle.

I had noted that: A few (singletons) have an obs admit with an inpatient LOS \> ER los (but not all).

Very likely valid approach: ignore the \~500 patients who are singletons (only one row in the ED Admissions table). Because so few have actual inpatient-looking encounters anyway.

`Pat Enc CSN` will be needed eventually to join up *notes* to *encounters.*

# Demographics (Table 1)

```{r tidy demographics, echo=FALSE}
demog %>%
  mutate(
    mcaid = ! is.na(MEDICAID_NUM),
    mcare = (MEDICARE_NUM != "")
  ) %>%
  select(
    pat_id, Sex, Age, Ethnic.Group, Race,
    Employment.Status,INTRPTR_NEEDED_YN,
    Language, Marital.Status, mcaid, mcare, Financial.Class
  ) %>%
  rename(
    interpreter = INTRPTR_NEEDED_YN,
  ) -> demographics_tidy

demographics_tidy %>%
  arrange(pat_id) %>%
  group_by(pat_id) %>%
  summarise(  
    demog_entries = n(),
    pat_id = first(pat_id),
    Sex = first(Sex),
    Age = first(Age),
    Ethnic.Group = first(Ethnic.Group),
    Race = first(Race),
    Employment.Status = first(Employment.Status),
    interpreter = first(interpreter),
    Language = first(Language),
    Marital.Status = first(Marital.Status),
    mcaid = first(mcaid),
    mcare = first(mcare),
    Financial.Class = first(Financial.Class)
  ) -> demographics_tidy_rm_dups
# fixme - probably a way to do w/ less copy/paste, like summarise_all(), etc.

```


**Observations in brief:** Number of patients is 3494. 2:1 white:black. Most common employment is retired. Most are married. 2:1 nonmedicare vs yes. Most common financial status is: managed care, or Medicare managed care.

```{r tableone}
  demographics_tidy_rm_dups %>% select(-pat_id) -> tabulate_me
  CreateTableOne(data=tabulate_me) -> t1
  kableone(t1)
```

```{r age density}
ggplot(demographics_tidy_rm_dups, aes(Age)) +   geom_density()
```

## Skip this section (Andy's notes)


```{r plot mult demog, echo=FALSE, message=FALSE, warning=FALSE}

demographics_tidy %>%
  group_by(pat_id) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) -> demog_rows_per_pt

table(demog_rows_per_pt$n) %>% kable()

```

```{r table mult demog, echo=FALSE}
left_join(demog_rows_per_pt, demographics_tidy,
          by="pat_id") -> mult_demog


mult_demog %>%
  filter(n == 4) %>%
  mutate(decade = round(Age, -1)) %>%
  select(-pat_id, -Age) %>%
  head(n = 12) -> inspect_mult_demog

# inspect_mult_demog %>%  kable()
```

**Table. People with multiple demographic entries.** Seems it is often for multiple financial classes. Sometimes for multiple races, though.



# Clinical Notes (descriptive stats)

```{r dim notes}
notes %>%
  group_by(PAT_ID) %>%
  summarise(note_count = n()) -> counted_notes

dim(counted_notes)
```

There are 100 patients for whom we have notes.

## Note chunks per patient

```{r message=FALSE, warning=FALSE}
ggplot(counted_notes, aes(note_count)) + geom_histogram() + scale_x_log10()
```

**Figure. Notes per patient.** About 35 patients have only one note chunk.

## Zoomed in note chunks per patient

```{r message=FALSE, warning=FALSE}
counted_notes %>%
  filter(note_count > 1) -> multi_notes
ggplot(multi_notes, aes(note_count)) + geom_histogram() + scale_x_log10()

```

**Figure. Zoomed in note chunks per patient.** Many people have dozens of note chunks. Plenty have hundreds. A few have thousands(!)

## Actual text processing

Review this: <https://pubmed.ncbi.nlm.nih.gov/35044842/> . Sun M, Oliwa T, Peek ME, Tung EL. Negative Patient Descriptors: Documenting Racial Bias In The Electronic Health Record. *Health Aff (Millwood).* 2022;41(2):203-211. <doi:10.1377/hlthaff.2021.01423>

> Fifteen descriptors were selected for inclusion in the analysis: (non-)adherent, aggressive, agitated, angry, challenging, combative, (non-)compliant, confront, (non-)cooperative, defensive, exaggerate, hysterical, (un-)pleasant, refuse, and resist. We adjusted the descriptors to permit identification of alternative grammatical forms (for example, "adher" for "adherent," "adhere," or "adhered)

> From all sentences in the data set, we selected a random sample of sentences containing one or more of the fifteen selected patient descriptors for manual review .... We categorized the use of each descriptor in one of three possible ways: negative, positive, or out of context.

> A total of 6,818 sentences were classified.

```{r text processing}
notes %>%
  mutate(negativity = count_descriptors(NOTE_TEXT)) %>%
  mutate(abbr = substr(NOTE_TEXT, 1, 100)) %>%
  select(PAT_ID, NOTE_ID, LINE, NOTE_TEXT, abbr, negativity) -> done_processing
```

```{r sum note lines}
done_processing %>%
  arrange(NOTE_ID, LINE) %>%
  group_by(NOTE_ID) %>%
  summarise(negativity = sum(negativity),
            PAT_ID = first(PAT_ID),
            NOTE_ID = first(NOTE_ID),
            abbr = first(abbr)
            ) -> summed
```

## Distribution of neg. descriptors per patient

```{r message=FALSE, warning=FALSE}
qplot(summed$negativity) + scale_y_log10()
```

**Figure. Negative descriptors per note.** There are 10k notes with 0 negative descriptors, 1k with 1, 100 with 3--4, etc.

# Final analytic dataset

```{r analytic dataset}
left_join(summed, demographics_tidy_rm_dups, by = c('PAT_ID' = 'pat_id')) %>%
  select(-abbr) -> joined
write.csv(joined, here('analytic_dataset.csv'))
names(joined)
```

```{r more concise analytic}
joined %>%
  mutate(negativity = (negativity > 0)) %>%
  filter(Race == 'White or Caucasian' |
           Race == 'Black or African American') -> analytic_concise
```


```{r samples, echo=FALSE}
joined %>%
  head() %>%
  write.csv(here('analytic_dataset_sample.csv'))

joined %>%
  select(-PAT_ID, -NOTE_ID, -mcaid, -mcare, -demog_entries, -Marital.Status) %>%
  rename(ethnic = Ethnic.Group,
         employ = Employment.Status,
         finance = Financial.Class,
         interp = interpreter,
         lang = Language
  ) %>%
  slice_sample(n=6) %>%
  kable()
```
**Table. Example analytic data set.** I printed the outcome variable
`negativity`, but only a selection of covariates, for simplicity. All covariates
are retained in the output CSV files.


# RESULTS: Negative descriptors vs. demographic factors

In these tables, an upward trend in ratio means positive correlation of underprivileged group status and negative descriptors. In other words, ratios are $r = u / p$, where $u$ is the count from a group hypothesized to be underprivileged, and $p$ is the group hypothesized to be privileged.

## Sex vs. negative descriptors

```{r}
table(joined$Sex, joined$negativity) -> x
round((x[1,] / x[2,]), 3) -> f.m.ratio
rbind(x, f.m.ratio) %>% kable()
```

**Table. Sex vs. negative descriptors.** More women when negativity = 1 or 2.

## Ethnicity vs. negative descriptors

```{r}
table(joined$Ethnic.Group, joined$negativity)  -> x
round((x[1,] / x[2,]), 3) -> latino.non.ratio
rbind(x, latino.non.ratio) %>% kable()
```

**Table. Latino ethnicity vs. negative descriptors.** More latino when negativity = 1 or 2.

## Race vs. negative descriptors

```{r}
table(joined$Race, joined$negativity)  -> x
round((x[2,] / x[5,]), 2) -> b.w.ratio  # warning, ad-hoc row numbers
rbind(x, b.w.ratio) %>% kable()
```

**Table. Race vs. negative descriptors.** Fewer AA when negativity = 1, more when = 2, 3, 4, 5, 6.

# RESULTS more concise

## Sex

```{r}
table(analytic_concise$Sex, analytic_concise$negativity) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion
rbind(x, proportion) %>% kable()
```


## Ethnicity

```{r}
table(analytic_concise$Ethnic.Group, analytic_concise$negativity)  -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion
rbind(x, proportion) %>% kable()
```


## Race

```{r}
table(analytic_concise$Race, analytic_concise$negativity)  -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion
rbind(x, proportion) %>% kable()
```



# To do

-  Make the proportion out of *all* patients. So $u / N$, instead of $u / (u + p)$.
-  Do stats like `prop.trend.test` / Cochran-Armitage test for trend.
-  Limit only to those discharged from ED to home.
-  Get complete note data from IT.
-  Consider showing which descriptor (arrange like doc:term matrix).
-  Show length distribution of notes.
-  Dunning log-likelihood.
-  Logistic model
