---
title: "BCM Datathon Team 17 Cleaning"
author: "Andrew Zimolzak, MD, MMSc"
date: 'April, 2022'
output:
  pdf_document: default
  html_document: default
---

# Setup
```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(dplyr)
library(lubridate)
library(data.table)
library(knitr)
library(tableone)
# library(readxl)

source(file=here("functions.R"))
```

Import data:

```{r import}

demog = filename2df('Patient Demographics 17 20220326.txt')  # 0.6 MB
# comor = filename2df('Patient Comorbidities 17 20220327.txt')  # 2.7 MB
edadm = filename2df('Patient ED Admissions 17 20220326.txt')  # 2.9 MB
# edpro = filename2df('Patient ED DX Problem List 17 20220327.txt')  # 19 MB
# medad = filename2df('Patient Meds Admin Times 17 20220327.txt')  # 93 MB
# order = filename2df('Patient Orders 17 20220327.txt')  # 86 MB
notes = read.csv(here('onedrive', 'Working', 'DATA', 'Progress Notes File.csv'),
                 sep=",", stringsAsFactors = FALSE)
```

Tidy up the data:

```{r tidy encounters}
edadm %>%
  mutate(
    ed_adm = ymd_hms(EMERGENCY_ADMISSION_DTTM),
	  ed_dis = ymd_hms(ED_DEPARTURE_DTTM),
	  inp_adm = ymd_hms(HOSP_ADMSN_TIME),
	  inp_dis = ymd_hms(HOSP_DISCH_TIME),
  ) %>%
  rename(cc10 = First.Chief.Compliant.ICD10, cc9 = First.Chief.Compliant.ICD9) %>%
  select(
	  PAT_ID, ED.Disposition, Patient.Class, cc10, cc9,
	  ed_adm, ed_dis, inp_adm, inp_dis
  ) %>%
  mutate(
	  adm_adm = (inp_adm - ed_adm) / ddays(),
	  dis_dis = (inp_dis - ed_dis) / ddays(),
	  ed_dur = (ed_dis - ed_adm) / ddays(),
	  inp_dur = (inp_dis - inp_adm) / ddays(),
	  er_vs_inp = inp_dur - ed_dur
  ) -> ed_encounter_tidy
```

# Descriptive tables

## Patient Class

```{r}
table(ed_encounter_tidy$Patient.Class) %>% kable()
```

## ED Disposition

```{r}
table(ed_encounter_tidy$ED.Disposition) %>% kable()
```

## Crosstabulation (disposition X class)

```{r}
table(ed_encounter_tidy$ED.Disposition, ed_encounter_tidy$Patient.Class) %>% kable()

```

Observations:

- disposition=Discharge and class=Emergency: common but not perfectly?
- d=Admit c=Inpatient: common
- d=Observation c=Observation: somewhat common
- d=Observation c=Inpatient: even more common (surprisingly)
- d=Admit class=Observation: common
- c=Outpatient: uncommon in general
- disposition=Admit class=Emergency: rare

The main dispositions that seem to matter:

- Discharge
- Admit
- Observation
- Transfer

We could make a more limited table of 4 dispo X 3 class. This is all possibly because we requested ER discharge to home and not admit?

# Sample data tables for inspection

Write a big one to disk:

```{r}
ed_encounter_tidy %>%
  select(-PAT_ID, -cc10, -cc9) -> sample_me

sample_me %>%
  slice_sample(n = 100) %>%
  arrange(ED.Disposition, Patient.Class) %>%
  fwrite(here("sample.csv"))
```

Write a short one to output document:

```{r}
sample_me %>%
  slice_sample(n=20) %>%
  arrange(Patient.Class, ED.Disposition) %>%
  transmute(
    class = tolower(substr(Patient.Class, 1, 3)),
    dispo = tolower(substr(ED.Disposition, 1, 3)),
    # er1 = ed_adm,
    # er2 = ed_dis,
    # in1 = inp_adm,
    # in2 = inp_dis,
    er1_to_inp1 = round(adm_adm, 2),
    er2_to_inp2 = round(dis_dis, 2),
    los_er = round(ed_dur, 2),
    los_inp = round(inp_dur, 2),
    diff_los = round(er_vs_inp, 2)
  ) %>%
  kable()
```

All numeric values are in **hours.**

Looks like if dispo = discharge, then ER departure time = Hospital discharge time.

# Distribution of rows per patient ID

```{r, message=FALSE}
ed_encounter_tidy %>%
  count(PAT_ID) -> pat_id_counts

qplot(x=n, data=pat_id_counts) +
  scale_x_log10()
```

# Who are the patients with multiple rows?

```{r}
left_join(ed_encounter_tidy, pat_id_counts, by='PAT_ID') %>%
  filter(Patient.Class != 'Outpatient') %>%
  filter(
    ED.Disposition == 'Admit' |
      ED.Disposition == 'Discharge' |
      ED.Disposition == 'Observation' |
      ED.Disposition == 'Transfer to Another Facility'
  ) %>%
  mutate(
    multiple = case_when(
      n > 1 ~ 'mult',
      TRUE ~ 'single'
    )
  ) -> ed_enc_w_counts
```

## Univariate tables

```{r}
table(ed_enc_w_counts$multiple, ed_enc_w_counts$Patient.Class) %>% kable()
```

```{r}
table(ed_enc_w_counts$multiple, ed_enc_w_counts$ED.Disposition) %>% kable()
```

```{r}
ed_enc_w_counts %>% filter(multiple == 'mult') -> m
ed_enc_w_counts %>% filter(multiple == 'single') -> s
```

## Crosstab limited to singles
```{r}
table(s$Patient.Class, s$ED.Disposition) %>% kable()
```

## Crosstab limited to multiples
```{r}
table(m$Patient.Class, m$ED.Disposition) %>% kable()
```

## Inspect multiples

```{r}
m %>% 
  select(PAT_ID, Patient.Class, ED.Disposition,
         ed_adm, ed_dis, inp_adm, inp_dis, n) %>%
  filter(n < 5) %>%
  arrange(PAT_ID, ed_adm) %>%
  head(25) %>%
  kable() -> inspect_multiples

# inspect_multiples
```

## Inspect singles

```{r}
s %>% 
  select(PAT_ID, Patient.Class, ED.Disposition,
         ed_adm, ed_dis, inp_adm, inp_dis, er_vs_inp, n) %>%
  arrange(PAT_ID, ed_adm) %>%
  head(25) %>%
  kable() -> inspect_singles

# inspect_singles
```

## Important note about "singleton" patients

2022-04-13 meeting notes.

ED admiss and ED discharge is literally when they came/went from ED. Whereas, inpatient admit/dis date and time
are about when they *first showed up to hospital* no matter what part of hospital. Many of the obs people will
expect to have inp discharge time which is *after* the ED disch time. But not *all.* Occasional data error or whatever.

Also some things come thru diff in Slicer vs Caboodle.

I had noted that: A few (singletons) have an obs admit with an inpatient LOS > ER los (but not all).

Very likely valid approach: ignore the ~500 patients who are singletons (only one row in the ED Admissions table).
Because so few have actual inpatient-looking encounters anyway.

`Pat Enc CSN` will be needed eventually to join up *notes* to *encounters.*




# Demographics

```{r tidy demographics}
demog %>%
  mutate(
    mcaid = ! is.na(MEDICAID_NUM),
    mcare = (MEDICARE_NUM != "")
  ) %>%
  select(
    pat_id, Sex, Age, Ethnic.Group, Race, Employment.Status,INTRPTR_NEEDED_YN,
    Language, Marital.Status, mcaid, mcare, Financial.Class
  ) %>%
  rename(
    interpreter = INTRPTR_NEEDED_YN,
  ) -> demographics_tidy

# demographics_tidy %>% head() %>% kable()
```

# Descriptives (of demographics)

```{r}
# addmargins(table(demographics_tidy$Sex)) %>% kable()
# table(demographics_tidy$Race) %>% kable()
# table(demographics_tidy$Ethnic.Group) %>% kable()
# table(demographics_tidy$Employment.Status) %>% kable()
# table(demographics_tidy$interpreter) %>% kable()
# table(demographics_tidy$Language) %>% kable()
# table(demographics_tidy$Marital.Status) %>% kable()
# table(demographics_tidy$mcaid) %>% kable()
# table(demographics_tidy$mcare) %>% kable()
# table(demographics_tidy$Financial.Class) %>% kable()
```

Sum is 3728. 2:1 white:black. Most common employment is retired. Most married.
2:1 nonmedicaid vs yes. Most common managed care and Medicare managed care.

```{r}
  demographics_tidy %>% select(-pat_id) -> tabulate_me
  CreateTableOne(data=tabulate_me) -> t1
  kableone(t1)
```

```{r}
ggplot(demographics_tidy, aes(Age)) +   geom_density()
```

# Notes

```{r}
notes %>%
  group_by(PAT_ID) %>%
  summarise(note_count = n()) -> counted_notes

dim(counted_notes)
```

There are 100 patients for whom we have notes.

```{r}
ggplot(counted_notes, aes(note_count)) + geom_histogram() + scale_x_log10()
```

About 35 have only one note.

```{r}
counted_notes %>%
  filter(note_count > 1) -> multi_notes
ggplot(multi_notes, aes(note_count)) + geom_histogram() + scale_x_log10()

```

Many people have dozens of notes. Plenty have hundreds. A few have thousands(!)

Review this: https://pubmed.ncbi.nlm.nih.gov/35044842/ . Sun M, Oliwa T, Peek ME, Tung EL. Negative Patient Descriptors: Documenting Racial Bias In The Electronic Health Record. *Health Aff (Millwood).* 2022;41(2):203-211. doi:10.1377/hlthaff.2021.01423

# Actual text processing

> Fifteen descriptors were selected for inclusion in the analysis: (non-)adherent, aggressive, agitated, angry, challenging, combative, (non-)compliant, confront, (non-)cooperative, defensive, exaggerate, hysterical, (un-)pleasant, refuse, and resist. We adjusted the descriptors to permit identification of alternative grammatical forms (for example, “adher” for “adherent,” “adhere,” or “adhered)

> From all sentences in the data set, we selected a random sample of sentences containing one or more of the fifteen selected patient descriptors for manual review .... We categorized the use of each descriptor in one of three possible ways: negative, positive, or out of context.

> A total of 6,818 sentences were classified.

```{r text processing}
notes %>%
  mutate(negativity = count_descriptors(NOTE_TEXT)) %>%
  mutate(abbr = substr(NOTE_TEXT, 1, 100)) %>%
  select(PAT_ID, NOTE_ID, NOTE_TEXT, abbr, negativity) -> done_searching
```

```{r}
qplot(done_searching$negativity) + scale_y_log10()
```

10k with 0, 1k with 1, 100 with 2, 10 with 3.

```{r}
left_join(done_searching, demographics_tidy, by = c('PAT_ID' = 'pat_id')) %>%
  select(-NOTE_TEXT, -abbr) -> joined
```

# Tables of negativity vs. various factors

```{r}
table(joined$Sex, joined$negativity) -> x
(x[1,] / x[2,]) -> f.m.ratio
rbind(x, f.m.ratio) %>% kable()
```
More women when negativity = 1 or 2.

```{r}
table(joined$Ethnic.Group, joined$negativity)  -> x
(x[2,] / x[1,]) -> non.latino.ratio
rbind(x, non.latino.ratio) %>% kable()
```
More latino when negativity = 1 or 2

```{r}
table(joined$Race, joined$negativity)  -> x
(x[3,] / x[6,]) -> b.w.ratio
rbind(x, b.w.ratio) %>% kable()
```
More AA when negativity = 1, fewer when = 2.

# To Do

1. Exclude stopwords (maybe n/a)
2. Join note chunks
3. Consider showing which descriptor (arrange like doc:term matrix)
4. Do stats like `prop.trend.test` / Cochran–Armitage test for trend
5. Make a separate pdf for presentation to team / large group
6. Show length distribution of notes
7. Dunning log-likelihood
