---
title: "Misc results, datathon 2022"
author: "Andrew Zimolzak"
date: '2022-04-20'
output: pdf_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(knitr)
library(dplyr)
library(ROCR)
```

```{r read CSV, echo=FALSE}
joined <- read.csv(
  here('analytic_dataset_tm.csv'), sep=",",
  stringsAsFactors = FALSE
) 
```




# Association results

The analysis is pretty sensitive to the phrasing of the descriptors. Limiting to "unpleasant" results in far fewer notes than "pleasant." Also, making that change results in more negativity for: *men, non-Latinos, and black patients,* unlike the prior word list when it was just the substring "pleasant."

In these tables, an upward trend in ratio means positive correlation of underprivileged group status and negative descriptors. In other words, ratios are $r = u / p$, where $u$ is the count from a group hypothesized to be underprivileged, and $p$ is the group hypothesized to be privileged.

## Sex vs. negative descriptors

```{r table s}
table(joined$Sex, joined$negativity) -> x
round((x[1,] / x[2,]), 3) -> f.m.ratio  # adhoc
rbind(x, f.m.ratio) %>% kable()
```

```{r trend test s}
women <- x[1,]  # adhoc
totals <- x[1,] + x[2,]  # adhoc
prop.trend.test(women, totals)
```

## Ethnicity vs. negative descriptors

```{r table e}
table(joined$Ethnic.Group, joined$negativity)  -> x
round((x[1,] / x[2,]), 3) -> latino.non.ratio  # adhoc
rbind(x, latino.non.ratio) %>% kable()
```

```{r trend test e}
latino <- x[1,]  # adhoc
totals <- x[1,] + x[2,]  # adhoc
prop.trend.test(latino, totals)
```

## Race vs. negative descriptors

```{r table race}
table(joined$Race, joined$negativity)  -> x
round((x['Black or African American',] / x['White or Caucasian',]), 2) -> b.w.ratio
rbind(x, b.w.ratio) %>% kable()
```

```{r trend test race}
black <- x['Black or African American',]
totals <- colSums(x)
prop.trend.test(black, totals)
```

## Single proportion line plot

```{r plot-multi-line, echo=FALSE, message=FALSE, warning=FALSE}
xvals = as.numeric(colnames(x))
N = length(xvals)

line_plot_me = data.frame(
  Proportion = c(women / totals, black / totals, latino / totals),
  Group = c(rep('Women', N), rep('Black', N), rep('Latino', N)),
  negativity = rep(xvals, 3)
)


ggplot(line_plot_me, aes(x=negativity, y=Proportion, color=Group)) +
  labs(x='Negative descriptors in note') +
  geom_line() +
  ylim(0,1)


```




# Combined race/ethnicity vs. negative descriptors

## Without a "cap" on negativity

```{r table combo, echo=FALSE}
table(joined$race_ethn, joined$negativity) -> x

Total <- colSums(x)
p_asian <- round(x['Asian',] / Total, 3)
p_black <- round(x['Black or African American',] / Total, 3)
p_latino <- round(x['Latino',] / Total, 3)
p_white <- round(x['White or Caucasian',] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table

unified_table %>% kable()
```

```{r trend-test-re-nocap-w}
prop.trend.test(x['White or Caucasian',], Total)
```

```{r trend-test-re-nocap-L}
prop.trend.test(x['Latino',], Total)
```

```{r trend-test-re-nocap-b}
prop.trend.test(x['Black or African American',], Total)
```

## With a "cap" on negativity

```{r table combo binned, echo=FALSE}
table(joined$race_ethn, joined$negativity_binned) -> x

Total <- colSums(x)
p_asian <- round(x['Asian',] / Total, 3)
p_black <- round(x['Black or African American',] / Total, 3)
p_latino <- round(x['Latino',] / Total, 3)
p_white <- round(x['White or Caucasian',] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table_binned

unified_table_binned %>% kable()
```

```{r areaplot, echo=FALSE, message=FALSE, warning=FALSE}
joined %>%
  select(race_ethn, negativity_binned) %>%
  group_by(race_ethn, negativity_binned) %>%
  summarise(tot = n()) %>%
  arrange(negativity_binned, race_ethn) -> race_neg_count

joined %>%
  group_by(negativity_binned) %>%
  summarise(grand_tot = n()) -> neg_count

left_join(neg_count, race_neg_count, by="negativity_binned") %>%
  mutate(proportion = tot / grand_tot) -> stack_plot_me

stack_plot_me %>%
  filter(race_ethn == 'Black or African American' & negativity_binned == 0) -> out
refaa <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'Latino' & negativity_binned == 0) -> out
refla <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'White or Caucasian' & negativity_binned == 0) -> out
refwh <- out$proportion

ggplot(stack_plot_me, aes(x = negativity_binned, y = proportion)) +
  geom_area(aes(fill = race_ethn)) +
  geom_hline(yintercept = refwh + refla) +
  geom_hline(yintercept = refwh)
```


```{r trend-test-re-w}
prop.trend.test(x['White or Caucasian',], Total)
```

```{r trend-test-re-L}
prop.trend.test(x['Latino',], Total)
```

```{r trend-test-re-b}
prop.trend.test(x['Black or African American',], Total)
```




# Logistic

## Big model

```{r echo=FALSE}
joined %>%
  select(
    -X, -pat_id, -NOTE_ID,
    -ends_with("_sum"), -n, -demog_entries,
    -negativity, -negativity_binned
  ) -> factor_me
  
factor_me %>%
  summarise(
    across(
      c(
        Sex, race_ethn, Employment.Status, interpreter, Language,
        Marital.Status, mcaid, mcare, Financial.Class
      ), factor
    ),
    Age = Age, negativity_any = negativity_any
  ) -> logit_me
```

```{r logistic}

my_model <- glm(negativity_any ~ ., data = logit_me, family = "binomial")
summary(my_model)

```

## ROC curve

```{r rocr}
rocr_pred = prediction(my_model$fitted.values, my_model$y)
rocr_perf <- performance(rocr_pred, measure = "tpr", x.measure = "fpr")
auc = performance(rocr_pred, measure = "auc")
auc@y.values
```

```{r plot-roc}
plot(rocr_perf, col=rainbow(10), print.cutoffs.at=c(0.01, 0.1))
```

## Little model

```{r little model}
little_model <- glm(negativity_any ~ Sex + race_ethn + Age, data = logit_me, family = "binomial")
summary(little_model)
```

## ROC curve

```{r}
rocr_pred = prediction(little_model$fitted.values, little_model$y)
rocr_perf <- performance(rocr_pred, measure = "tpr", x.measure = "fpr")
auc = performance(rocr_pred, measure = "auc")
auc@y.values
```

```{r}
plot(rocr_perf, col=rainbow(10), print.cutoffs.at=c(0.01, 0.1))
```

## Predicted probabilities, big vs. little

```{r message=FALSE, warning=FALSE}
qplot(my_model$fitted.values) + labs(title = 'Big model')
```

```{r message=FALSE, warning=FALSE}
qplot(little_model$fitted.values) + labs(title = 'Little model') + xlim(0, 0.15)
```

I get the feeling it hates unbalanced classes.

## Try column plot of coefficients

```{r}
df = data.frame(
  y=my_model$coefficients,
  x=names(my_model$coefficients)
)

ggplot(df, aes(x=y, y=x) ) + geom_col() + scale_x_log10() +
  geom_vline(xintercept = 1)

```




# To do, or not

-  Limit only to those discharged from ED to home.
-  More rigorous handling of those with multiple demographic entries.
-  Get complete note data from IT.
-  Show length distribution of notes.

