---
title: "BCM Datathon Team 17 Results"
author: "Andrew Zimolzak, MD, MMSc"
date: 'April, 2022'
output: pdf_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(knitr)
library(dplyr)
```

# RESULTS: Negative descriptors vs. demographic factors

Determine which data set we're using. `joined` is the first one I made, which counts how many of our ~16 terms occur in each note chunk, and then adds that up. `joined_tm` is the more pre-packaged method. Counts the *number of mentions* of each term, per note.

```{r}
joined <- read.csv(
  here('analytic_dataset_tm.csv'), sep=",",
  stringsAsFactors = FALSE
) 
```


In these tables, an upward trend in ratio means positive correlation of underprivileged group status and negative descriptors. In other words, ratios are $r = u / p$, where $u$ is the count from a group hypothesized to be underprivileged, and $p$ is the group hypothesized to be privileged.

## Sex vs. negative descriptors

```{r table s}
table(joined$Sex, joined$negativity) -> x
round((x[1,] / x[2,]), 3) -> f.m.ratio
rbind(x, f.m.ratio) %>% kable()
```

**Table. Sex vs. negative descriptors.** More women when negativity = 1 or 2.

```{r trend test s}
women <- x[1,]
totals <- x[1,] + x[2,]
prop.trend.test(women, totals)
```

```{r plot-s, echo=FALSE, message=FALSE, warning=FALSE}
xvals = as.numeric(colnames(x))
p_women = women / totals

qplot(x = xvals, y = p_women) +
  labs(y='Proportion Women', x='Negative descriptors in note') +
  geom_line() +
  geom_hline(yintercept = p_women[1])
```

## Ethnicity vs. negative descriptors

```{r table e}
table(joined$Ethnic.Group, joined$negativity)  -> x
round((x[1,] / x[2,]), 3) -> latino.non.ratio
rbind(x, latino.non.ratio) %>% kable()
```

**Table. Latino ethnicity vs. negative descriptors.** More latino when negativity = 1 or 2.

```{r trend test e}
latino <- x[1,]
totals <- x[1,] + x[2,]
prop.trend.test(latino, totals)
```

```{r plot-e, echo=FALSE, message=FALSE, warning=FALSE}
xvals = as.numeric(colnames(x))
p_latino = latino / totals

qplot(x = xvals, y = p_latino) +
  labs(y='Proportion Latino', x='Negative descriptors in note') +
  geom_line() +
  geom_hline(yintercept = p_latino[1])
```


## Race vs. negative descriptors

```{r table}
table(joined$Race, joined$negativity)  -> x
round((x[2,] / x[5,]), 2) -> b.w.ratio  # warning, ad-hoc row numbers
rbind(x, b.w.ratio) %>% kable()
```

**Table. Race vs. negative descriptors.** Fewer AA when negativity = 1, more when = 2, 3, 4, 5, 6.

```{r trend test}
black <- x[2,]  # warning - ad hoc row numbers again
totals <- colSums(x)
prop.trend.test(black, totals)
```

```{r plot, echo=FALSE, message=FALSE, warning=FALSE}
xvals = as.numeric(colnames(x))
p_black = black / totals

qplot(x = xvals, y = p_black) +
  labs(y='Proportion Black', x='Negative descriptors in note') +
  geom_line() +
  geom_hline(yintercept = p_black[1])
```

## Combined race/ethnicity vs. negative descriptors

```{r table combo, echo=FALSE}
table(joined$race_ethn, joined$negativity) -> x

Total <- colSums(x)
p_asian <- round(x[1,] / Total, 3)
p_black <- round(x[2,] / Total, 3)
p_latino <- round(x[3,] / Total, 3)
p_white <- round(x[4,] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table

unified_table %>% kable()
```

```{r table combo binned, echo=FALSE}
table(joined$race_ethn, joined$negativity_binned) -> x

Total <- colSums(x)
p_asian <- round(x[1,] / Total, 3)
p_black <- round(x[2,] / Total, 3)
p_latino <- round(x[3,] / Total, 3)
p_white <- round(x[4,] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table_binned

unified_table_binned %>% kable()
```


```{r echo=FALSE}
g <- ggplot(joined, aes(negativity_binned))
g + geom_bar(aes(fill = race_ethn)) + scale_y_log10()
```

```{r create-interesting, echo=FALSE, message=FALSE, warning=FALSE}
joined %>%
  select(race_ethn, negativity_binned) %>%
  group_by(race_ethn, negativity_binned) %>%
  summarise(tot = n()) %>%
  arrange(negativity_binned, race_ethn) -> race_neg_count

joined %>%
  group_by(negativity_binned) %>%
  summarise(grand_tot = n()) -> neg_count

left_join(neg_count, race_neg_count, by="negativity_binned") %>%
  mutate(proportion = tot / grand_tot) -> stack_plot_me

stack_plot_me %>%
  filter(race_ethn == 'Black or African American' & negativity_binned == 0) -> out
refaa <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'Latino' & negativity_binned == 0) -> out
refla <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'White or Caucasian' & negativity_binned == 0) -> out
refwh <- out$proportion

ggplot(stack_plot_me, aes(x = negativity_binned, y = proportion)) +
  geom_area(aes(fill = race_ethn)) +
  geom_hline(yintercept = refwh + refla) +
  geom_hline(yintercept = refwh)
```


# RESULTS more concise

## Sex

```{r}
table(joined$Sex, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion
rbind(x, proportion) %>% kable()
```


## Ethnicity

```{r}
table(joined$Ethnic.Group, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion
rbind(x, proportion) %>% kable()
```


## Race

```{r}
table(joined$Race, joined$negativity_any) -> x
colSums(x) -> total
round((x[2,] / total), 3) -> proportion.black
rbind(x, total, proportion.black) %>% kable()
```

# Skip this section (R vs E)

```{r}
table(joined$Race, joined$Ethnic.Group)
```


# To do

-  cut out the "skip this" sections
-  echo less code
-  All dot/line plots on one set of axes
-  Logistic model
-  Dunning log-likelihood.
-  Second look at `descrip`. "pleasant" vs "unpleasant", "refuse" vs "refus".
-  Limit only to those discharged from ED to home.
-  More rigorous handling of those with multiple demographic entries.
-  Get complete note data from IT.
-  Consider showing which descriptor (arrange like doc:term matrix).
-  Show length distribution of notes.
