---
title: "The association of demographic factors with negative desriptors in medical text"
author: "Andrew Zimolzak, Traber Giardina, Darius Dawson, Terri Fletcher, Taylor Scott, Debra Choi"
date: 'April, 2022'
output: pdf_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(knitr)
library(dplyr)
```

```{r read CSV, echo=FALSE}
joined_initial <- read.csv(
  here('analytic_dataset_tm.csv'), sep=",",
  stringsAsFactors = FALSE
) 

joined_initial %>%
  rename(negativity_bool = negativity_any) %>%
  mutate(
    negativity_any = case_when(
      negativity_bool == TRUE ~ 'Has negative descriptors',
      negativity_bool == FALSE ~ 'No negative descriptors',
    )
  ) -> joined
```

```{r write-cleaner-csv, echo=FALSE}
joined %>%
  select(-ends_with("_sum"), -n, -demog_entries) -> out

out %>%
  write.csv(here('analytic dataset v0xx 2022-04-xx.csv'))
```

# Introduction

Recent work used an expert panel to develop a list of negative patient
descriptors and found that Black patients had disproportionately
higher odds of negative patient descriptors appearing in the history
and physical notes of their medical records compared with White
patients. Sun *et al.* used the following 15 descriptors:

> (non-)adherent, aggressive, agitated, angry, challenging, combative,
> (non-)compliant, confront, (non-)cooperative, defensive, exaggerate,
> hysterical, (un-)pleasant, refuse, and resist.

> **Source:** Sun M, Oliwa T, Peek ME, Tung EL. Negative Patient Descriptors:
> Documenting Racial Bias In The Electronic Health Record.
> *Health Aff (Millwood).* 2022 Feb;41(2):203-211.
> PMID: [35044842](https://pubmed.ncbi.nlm.nih.gov/35044842/).

We measured the association of race, ethnicity, and sex with these
negative descriptor words (and word variants), in the clinical notes of 100 patients who
had an ER visit and inpatient stay.

# Methods

Data cleaning, analysis, and reporting was performed using R Markdown software.
All code (including code used to generate this report) can be found at
https://github.com/zimolzak/datathon-2022 . There were 115 edits to the code
from April 5 to April 20 (average 7.7 edits per day). As of April 20, the project
comprises about 1000 lines of code (average 68 lines per day).




# Tables and figures

```{r fig-area, echo=FALSE, message=FALSE, warning=FALSE}
joined %>%
  select(race_ethn, negativity_any) %>%
  group_by(race_ethn, negativity_any) %>%
  summarise(tot = n()) %>%
  arrange(negativity_any, race_ethn) -> race_neg_count

joined %>%
  group_by(negativity_any) %>%
  summarise(grand_tot = n()) -> neg_count

left_join(neg_count, race_neg_count, by="negativity_any") %>%
  mutate(Proportion = tot / grand_tot) -> stack_plot_me

ggplot(stack_plot_me, aes(x=negativity_any, y=Proportion)) +
  geom_col(aes(fill = race_ethn)) +
  labs(x="") +
  scale_fill_discrete("Race/ethnicity") -> final_plot

final_plot
```

**Figure. Distribution of race/ethnicity, in notes with negative
descriptors, compared to notes without negative descriptors.** Latino
patients are overrepresented among notes that contain one or more
negative descriptor words. Here, the Latino grouping comprises *any*
race, whereas the other groupings comprise *non-Latino* patients of
the specified race.




```{r table-race-ethn, echo=FALSE}
table(joined$race_ethn, joined$negativity_any) -> final_race
fisher.test(final_race)[["p.value"]] -> final_race_p
final_race %>% kable()
```

**Table 1. Distribution of race/ethnicity, in notes with negative
descriptors, compared to notes without negative descriptors.**
The population with negative descriptors is significantly different from the
population without negative descriptors (in terms of race/ethnicity distribution).
*P* = `r final_race_p`, Fisher's exact test.




```{r table-sex, echo=FALSE}
table(joined$Sex, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> Proportion.women  # adhoc
fisher.test(x)[["p.value"]] -> sex_p
rbind(x, Proportion.women) %>% kable()
```

**Table 2. Distribution of sex, in notes with/without negative
descriptors.** There are proportionally fewer women / more men
represented among notes containing negative descriptors.
*P* = `r sex_p`




```{r table-ethn, echo=FALSE}
table(joined$Ethnic.Group, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> Proportion.Latino  # adhoc
fisher.test(x)[["p.value"]] -> ethn_p
rbind(x, Proportion.Latino) %>% kable()
```

**Table 3. Distribution of ethnicity, in notes with/without negative
descriptors.** There are proportionally more Latino patients
represented among notes containing negative descriptors.
*P* = `r ethn_p`





```{r table-race, echo=FALSE}
table(joined$Race, joined$negativity_any) -> x
colSums(x) -> Total
round((x['Black or African American',] / Total), 3) -> Proportion.black
fisher.test(x)[['p.value']] -> race_alone_p
rbind(x, Total, Proportion.black) %>% kable()
```

**Table 4. Distribution of race, in notes with/without negative
descriptors.** There are more black / African-American patients
represented among notes containing negative descriptors. Here, Latino
patients can be in any grouping, although the majority are white
Latino. *P* = `r race_alone_p`




# Discussion

```{r math-for-discussion, echo=FALSE}
sig_figs = 3
ps = signif(sex_p, sig_figs)
pe = signif(ethn_p, sig_figs)
pr = signif(race_alone_p, sig_figs)
pc = signif(final_race_p, sig_figs)
```

There are significantly **more Latino patients**
(`r Proportion.Latino[2]*100` vs. `r Proportion.Latino[1]*100` percent)
than non-Latinos represented among notes containing any negative
descriptor (*P* = `r pe`). There are significantly **more black
patients** (`r Proportion.black[2]*100` vs. 
`r Proportion.black[1]*100` percent) represented among notes with any
negative descriptor (*P* = `r pr`). There are significantly **fewer
women** (`r Proportion.women[2]*100` vs. `r Proportion.women[1]*100`
percent) represented among notes containing any negative descriptor,
compared to notes without a negative descriptor (*P* = `r ps`).
Considering the *combination* of race and ethnicity fields, (that is,
with the major groups being black, Latino, and white non-Latino,)
there is a significant difference in the overall distribution of groups,
between notes with vs. without a negative descriptor (*P* = `r pc`).

Limitation: Negative descriptors are really only *potentially* negative descriptors.
Sun *et al.* were able to classify usages as negative, positive, or neutral.
We have not implemented that technique.




# Data lessons learned

- Data retrieval and cleaning takes time!
- Handing off a dataset is not trivial (receiving end does not know how it was made, or column meanings, and may need explanations for unexpected associations).
- Project management: team members often work in tandem, not in parallel.
- Sometimes challenging to involve all team members.
- It is possible to do things fast.
- Record keeping is important, even when going fast.
- With "secondary use," researchers make do with the available data. The warehouse may not have every data element needed to answer certain questions.
