---
title: "BCM Datathon Team 17 Results"
author: "Andrew Zimolzak, MD, MMSc"
date: 'April, 2022'
output: pdf_document
---

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
library(knitr)
library(dplyr)

source(file=here("functions.R"))
```

# Conclusions so far

There are more women and Latino patients represented among notes containing any negative descriptor, compared to notes without a negative descriptor. There are slightly fewer black patients represented among notes with any negative descriptor.

Pretty sensitive to phrasing of descriptors. "Unpleasant" results in far fewer notes. Also it means more negativity for: *men, non-Latinos, and black patients,* unlike the prior word list when it was just the substring "pleasant."

No adjustment yet for covariates. To see quantitative association results, go to "RESULTS" section (probably the final 4-8 pages of the PDF).



# RESULTS: Negative descriptors vs. demographic factors

Determine which data set we're using. `analytic_dataset.csv` is the first one I made, which counts how many of our ~16 terms occur in each note chunk, and then adds that up. `analytic_dataset_tm.csv` is the more pre-packaged method. Counts the *number of mentions* of each term, per note.

```{r}
joined <- read.csv(
  here('analytic_dataset_tm.csv'), sep=",",
  stringsAsFactors = FALSE
) 
```

In these tables, an upward trend in ratio means positive correlation of underprivileged group status and negative descriptors. In other words, ratios are $r = u / p$, where $u$ is the count from a group hypothesized to be underprivileged, and $p$ is the group hypothesized to be privileged.

## Our list of descriptors

```{r}
descrip
```


## Sex vs. negative descriptors

```{r table s}
table(joined$Sex, joined$negativity) -> x
round((x[1,] / x[2,]), 3) -> f.m.ratio  # adhoc
rbind(x, f.m.ratio) %>% kable()
```

```{r trend test s}
women <- x[1,]  # adhoc
totals <- x[1,] + x[2,]  # adhoc
prop.trend.test(women, totals)
```



## Ethnicity vs. negative descriptors

```{r table e}
table(joined$Ethnic.Group, joined$negativity)  -> x
round((x[1,] / x[2,]), 3) -> latino.non.ratio  # adhoc
rbind(x, latino.non.ratio) %>% kable()
```

```{r trend test e}
latino <- x[1,]  # adhoc
totals <- x[1,] + x[2,]  # adhoc
prop.trend.test(latino, totals)
```




## Race vs. negative descriptors

```{r table race}
table(joined$Race, joined$negativity)  -> x
round((x['Black or African American',] / x['White or Caucasian',]), 2) -> b.w.ratio
rbind(x, b.w.ratio) %>% kable()
```

```{r trend test race}
black <- x['Black or African American',]
totals <- colSums(x)
prop.trend.test(black, totals)
```



## Single proportion line plot

```{r plot-multi-line, echo=FALSE, message=FALSE, warning=FALSE}
xvals = as.numeric(colnames(x))
N = length(xvals)

line_plot_me = data.frame(
  Proportion = c(women / totals, black / totals, latino / totals),
  Group = c(rep('Women', N), rep('Black', N), rep('Latino', N)),
  negativity = rep(xvals, 3)
)


ggplot(line_plot_me, aes(x=negativity, y=Proportion, color=Group)) +
  labs(x='Negative descriptors in note') +
  geom_line() +
  ylim(0,1)


```

## Combined race/ethnicity vs. negative descriptors

*Without* a "cap" on negativity:

```{r table combo, echo=FALSE}
table(joined$race_ethn, joined$negativity) -> x

Total <- colSums(x)
p_asian <- round(x['Asian',] / Total, 3)
p_black <- round(x['Black or African American',] / Total, 3)
p_latino <- round(x['Latino',] / Total, 3)
p_white <- round(x['White or Caucasian',] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table

unified_table %>% kable()
```

*With* a "cap" on negativity:

```{r table combo binned, echo=FALSE}
table(joined$race_ethn, joined$negativity_binned) -> x

Total <- colSums(x)
p_asian <- round(x['Asian',] / Total, 3)
p_black <- round(x['Black or African American',] / Total, 3)
p_latino <- round(x['Latino',] / Total, 3)
p_white <- round(x['White or Caucasian',] / Total, 3)

rbind(x, Total, p_asian, p_black, p_latino, p_white) -> unified_table_binned

unified_table_binned %>% kable()
```



```{r areaplot, echo=FALSE, message=FALSE, warning=FALSE}
joined %>%
  select(race_ethn, negativity_binned) %>%
  group_by(race_ethn, negativity_binned) %>%
  summarise(tot = n()) %>%
  arrange(negativity_binned, race_ethn) -> race_neg_count

joined %>%
  group_by(negativity_binned) %>%
  summarise(grand_tot = n()) -> neg_count

left_join(neg_count, race_neg_count, by="negativity_binned") %>%
  mutate(proportion = tot / grand_tot) -> stack_plot_me

stack_plot_me %>%
  filter(race_ethn == 'Black or African American' & negativity_binned == 0) -> out
refaa <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'Latino' & negativity_binned == 0) -> out
refla <- out$proportion

stack_plot_me %>%
  filter(race_ethn == 'White or Caucasian' & negativity_binned == 0) -> out
refwh <- out$proportion

ggplot(stack_plot_me, aes(x = negativity_binned, y = proportion)) +
  geom_area(aes(fill = race_ethn)) +
  geom_hline(yintercept = refwh + refla) +
  geom_hline(yintercept = refwh)
```

```{r areaplot-any, echo=FALSE, message=FALSE, warning=FALSE}
joined %>%
  select(race_ethn, negativity_any) %>%
  group_by(race_ethn, negativity_any) %>%
  summarise(tot = n()) %>%
  arrange(negativity_any, race_ethn) -> race_neg_count

joined %>%
  group_by(negativity_any) %>%
  summarise(grand_tot = n()) -> neg_count

left_join(neg_count, race_neg_count, by="negativity_any") %>%
  mutate(proportion = tot / grand_tot) -> stack_plot_me


ggplot(stack_plot_me, aes(x=negativity_any, y=proportion)) + geom_col(aes(fill = race_ethn))
```


# RESULTS more concise

## Sex

```{r}
table(joined$Sex, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion  # adhoc
rbind(x, proportion) %>% kable()
```


## Ethnicity

```{r}
table(joined$Ethnic.Group, joined$negativity_any) -> x
round((x[1,] / (x[1,] + x[2,])), 3) -> proportion  # adhoc
rbind(x, proportion) %>% kable()
```


## Race

```{r}
table(joined$Race, joined$negativity_any) -> x
colSums(x) -> total
round((x['Black or African American',] / total), 3) -> proportion.black
rbind(x, total, proportion.black) %>% kable()
```

# Skip this section (R vs E)

```{r}
table(joined$Race, joined$Ethnic.Group)
```


# Most common negative descriptors


```{r}
joined %>%
  select(ends_with('_sum')) %>%
  summarise_all(~ sum(.))
```
Challenging = 69, angry = 12, total notes flagged = 80.

# To do

-  Trend test on 3 rows of the "combined" `race_ethn` table.
-  Second look at `descrip`. "pleasant" vs "unpleasant", "refuse" vs "refus".
-  Logistic model
-  Dunning log-likelihood.
-  Limit only to those discharged from ED to home.
-  More rigorous handling of those with multiple demographic entries.
-  Get complete note data from IT.
-  Show length distribution of notes.
